\chapter{Class 1}

\newcommand{\Reals}{\mathbb{R}}

\section{Tools}
\begin{itemize}
  \item Lean or Coq 
  \item CVC5 or Z3 
  \item possibly a model checker 
  \item professional version of ChatGPT
\end{itemize}

\section{Syllabus}
\begin{enumerate}
  \item MATH (``Informalism'')
  \begin{itemize}
    \item proofs (natural deduction)
    \item fixpoints (induction, coinduction)
  \end{itemize}
  \item DECLARATIVE LOGIC
  \begin{itemize}
    \item syntax (rules) vs. semantics (models)
    \item propositional, predicate, modal logic 
    \item decision procedures (SAT, SMT)
  \end{itemize}
  \item FUNCTIONS
  \begin{itemize}
    \item $\lambda$ calculus, typed $\lambda$
    \item SOS (structured operational semantics), rewriting
    \item ``propositions-as-types'' (connection to logic)
  \end{itemize}
  \item PROCESSES (CONCURRENT)* 
  \begin{itemize}
    \item CCS, Petri nets
    \item (bi-) simulation
  \end{itemize}
  \item CIRCUITS*
  \begin{itemize}
    \item boolean, sequential, dataflow (Kahn nets)
    \item interfaces
  \end{itemize}
  \item STATE TRANSITION SYSTEMS*
  \begin{itemize}
    \item ($\omega$-) automata, games, timed, probabilistic, pushdown
    \item programs, Turing machines 
    \item grammars (Chomsky hierarchy)
  \end{itemize} 
  \item DECLARATIVE SPECIFICATION
  \begin{itemize}
    \item Hoare logics, separation logic 
    \item temporal logics (LTL, CTL, ATL)
    \item partial correctness vs. termination, safety vs. liveness 
  \end{itemize}
\end{enumerate}

*: Operational.

\section{Math}
\begin{definition}
  A real $b$ is a \underline{bound} of a 
  function $f$ from $\Reals$ to $\Reals$ if for all $x$ in 
  $\Reals$, we have $f(x) \leq b$. 
\end{definition}

\begin{definition}
  Given two functions $f$ and $g$ from $\Reals$ to 
  $\Reals$, their \underline{sum} is the function $f + g$ such 
  that for all $x$ in $\Reals$, we have $(f+g) (x) = f(x) + g(x)$.
\end{definition} 

\begin{theorem}
  For all functions $f$ and $g$ from $\Reals$ to $\Reals$, if $f$ and
  $g$ are bounded, then $f+g$ is bounded. 
\end{theorem}
\begin{proof}
  \begin{enumerate}
    \item Consider arbitrary functions $\hat{f}$ and $\hat{g}$ from 
    $\Reals$ to $\Reals$.
    \item Assume $\hat{f}$ and $\hat{g}$ are bounded. 
    \item Show that $\hat{f} + \hat{g}$ is bounded.
    \item (2 $\rightarrow$) Let $\hat{a}$ be a bound for $\hat{f}$, 
    and $\hat{b}$ be a bound for $\hat{g}$.
    \item We show that $\hat{a} + \hat{b}$ is a bound for 
    $\hat{f} + \hat{g}$.
    \item Consider an arbitrary real $\hat{x}$.
    \item Show $(\hat{f}+\hat{g})(x) \leq \hat{a} + \hat{b}$.
    \item (Definition of sum) $(\hat{f}+\hat{g})(\hat{x}) 
    = \hat{f}(\hat{x}) + \hat{g}(\hat{x})$.
    \item (Definition of bound) $\hat{f}(\hat{x}) \leq \hat{a}$ and 
    $\hat{g}(\hat{x}) \leq \hat{b}$.
    \item The rest follows from ``arithmetic''. 
  \end{enumerate}
\end{proof}

\noindent \textbf{Homework.} Prove the Schr\"{o}der-Bernstein theorem 
``in this style''.

\begin{definition}
  Two sets $A$ and $B$ are \underline{equipollent} (``have the same 
  size'') if there is a bijection from $A$ to $B$.
\end{definition}

\begin{definition}
  A function $f$ from $A$ to $B$ is 
  \begin{enumerate}
    \item \underline{one-to-one} if for all $x$ and $y$ in $A$, if 
    $x \neq y$, then $f(x) \neq f(y)$.
    \item \underline{onto} if for all $z$ in $B$, there exists $x$ 
    in $A$ such that $f(x) = z$.
    \item \underline{bijective} if $f$ is one-to-one and onto.
  \end{enumerate}
\end{definition}

\begin{center}
\begin{tabular}{
  p{0.35\linewidth} | p{0.35\linewidth} | p{0.2\linewidth}}
  Goals & Knowledge & Outermost symbol \\
  \hline
  \begin{itemize}[leftmargin=*]
    \item[] Show for all $x$, $G(x)$.
    \item[] Consider arbitrary $\hat{x}$.
    \item[] Show $G(\hat{x})$. 
  \end{itemize} & \begin{itemize}[leftmargin=*]
    \item[] We know for all $x$, $K(x)$.
    \item[] In particular, we know $K(\hat{t})$.
    \item[] $\hat{t}$: term containing only constants. 
  \end{itemize} & \begin{itemize}
    \item[] \Large $\forall$
  \end{itemize} \\ 
  \hline 
  \begin{itemize}[leftmargin=*]
    \item[] Show there exists $x$ s.t. $G(x)$.
    \item[] We show that $G(\hat{t})$.
    \item[] $\hat{t}$: term containing only constants.
  \end{itemize} & \begin{itemize}[leftmargin=*]
    \item[] We know there exists $x$ s.t. $K(x)$.
    \item[] Let $\hat{x}$ be s.t. $K(\hat{x})$.
  \end{itemize} & \begin{itemize}
    \item[] \Large $\exists$
  \end{itemize}
\end{tabular}
\end{center}